---
title: "HW 2: From correlation to linear mixed-effect models. Assignment sheet"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(vcd)
library(lfe)
```

## 1. Vowel reduction in Russian
Pavel Duryagin ran an experiment on perception of vowel reduction in Russian language. The dataset `shva` includes the following variables:  
_time1_ - reaction time 1  
_duration_ - duration of the vowel in the stimuly (in milliseconds, ms)  
_time2_ - reaction time 2  
_f1_, _f2_, _f3_ - the 1st, 2nd and 3rd formant of the vowel measured in Hz (for a short introduction into formants, see [here](https://home.cc.umanitoba.ca/~krussll/phonetics/acoustic/formants.html))  
_vowel_ - vowel classified according the 3-fold classification (_A_ - _a_ under stress, _a_ - _a/o_ as in the first syllable before the stressed one, _y_ (stands for shva) - _a/o_ as in the second etc. syllable before the stressed one or after the stressed syllable, cf. _g_[_y_]_g_[_a_]_t_[_A_]_l_[_y_] _gogotala_ `guffawed').  
In this part, we will ask you to analyse correlation between f1, f2, and duration.
The dataset is available [https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt](here).

### 1.0 Read the data from file to the variable `shva`.
```{r 1.0}
shva <- read.csv('https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt', sep = '\t')
head(shva)
```

### 1.1 Scatterplot `f1` and `f2` using `ggplot()`. 
Design it to look like the [following](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin1.png).
```{r 1.1}
shva %>%
ggplot(aes(x = f2, y = f1, colour = vowel)) + geom_point() + labs(title = 'f2 and f1 of the reduced and stressed vowels', caption = 'Data from Duryagin 2018') + theme(legend.position = 'none') + scale_x_reverse() + scale_y_reverse()
```

### 1.2 Plot the boxplots of `f1` and `f2` for each vowel using `ggplot()`.
Design it to look like [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin2.png) and [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin3.png).
```{r 1.2}
# the 1st boxplot
shva %>%
ggplot(aes(x = vowel, y = f1, fill = vowel)) + geom_boxplot() + labs(title = 'f1 distribution in each vowel', caption = 'Data from Duryagin 2018') +  theme(legend.position = 'none') + coord_flip()
# the 2nd boxplot
shva %>%
ggplot(aes(x = vowel, y = f2, fill = vowel)) + geom_boxplot() + labs(title = 'f2 distribution in each vowel', caption = 'Data from Duryagin 2018') +  theme(legend.position = 'none') + coord_flip()
```

### 1.3 Which `f1` can be considered outliers in _a_ vowel?
We assume outliers to be those observations that lie outside 1.5 * IQR, where IQR, the 'Inter Quartile Range', is the difference between the 1st and the 3rd quartile (= 25% and 75% percentile).
```{r 1.3}
outliers <- subset(shva$f1, shva$vowel == 'a')
threshold <- 1.5 * IQR(outliers)
outliers_result <- subset(outliers, outliers < quantile(outliers)[2] - threshold | outliers > quantile(outliers)[4]  + threshold)
outliers_result
### >The f1 of 679 can be cosidered outliers.
```

### 1.4 Calculate Pearson's correlation of `f1` and `f2` (all data)
```{r 1.4}
cor(shva$f1, shva$f2, method = 'pearson')
```

### 1.5 Calculate Pearson's correlation of `f1` and `f2` for each vowel
```{r 1.5}
A_f1 <-subset(shva$f1,shva$vowel=='A')
A_f2 <-subset(shva$f2,shva$vowel=='A')
A_f1_f2_correlation <- cor(A_f1, A_f2, method = 'pearson')
a_f1 <-subset(shva$f1,shva$vowel=='a')
a_f2 <-subset(shva$f2,shva$vowel=='a')
a_f1_f2_correlation <- cor(a_f1, a_f2, method = 'pearson')
y_f1 <-subset(shva$f1,shva$vowel=='y')
y_f2 <-subset(shva$f2,shva$vowel=='y')
y_f1_f2_correlation <- cor(y_f1, y_f2, method = 'pearson')

A_f1_f2_correlation
a_f1_f2_correlation
y_f1_f2_correlation

```
###1.6 Use the linear regression model to predict f2 by f1.
###1.6.1 Provide the result regression formula
```{r 1.6.1}
LinearMod <- lm(f2 ~ f1, shva)
summary(LinearMod)
```
###1.6.2 Provide the adjusted R2
```{r 1.6.2}
###The adjusted R2 is equal to 0.3319.
```
###1.6.3 Add the regression line in scatterplot 1.1
```{r 1.6.3}
shva %>%
ggplot(aes(x = f2, y = f1, colour = vowel)) + geom_point() + labs(title = 'f2 and f1 of the reduced and stressed vowels', caption = 'Data from Duryagin 2018') + theme(legend.position = 'none') + scale_x_reverse() + scale_y_reverse() + geom_smooth(method = "lm", se = FALSE, colour = 'grey')
```
###1.7 Use the mixed-efects model to predict f2 by f1 using vowel intercept as a random effect
###1.7.1 Provide the fixed effects formula
```{r 1.7.1}
fixed_effects <- lmer(f2 ~ f1 + (1|vowel), shva)
summary(fixed_effects)
```
###1.7.2 Provide the variance for intercept argument for vowel random effects
```{r 1.7.2}
###The variance is equal to 6374.
```
###1.7.3 Add the regression line in scatterplot 1.1
```{r 1.7.3}
shva %>%
ggplot(aes(x = f2, y = f1, colour = vowel)) + geom_point() + labs(title = 'f2 and f1 of the reduced and stressed vowels', caption = 'Data from Duryagin 2018') + theme(legend.position = 'none') + scale_x_reverse() + scale_y_reverse() + geom_line(aes(y=fitted(fixed_effects), colour = shva$vowel))
```

## 2. English Lexicon Project data
880 nouns, adjectives and verbs from the English Lexicon Project data (Balota et al. 2007).

* `Format` -- A data frame with 880 observations on the following 5 variables.
* `Word` -- a factor with lexical stimuli.
* `Length` -- a numeric vector with word lengths.
* `SUBTLWF` -- a numeric vector with frequencies in film subtitles.
* `POS` -- a factor with levels JJ (adjective) NN (noun) VB (verb)
* `Mean_RT` -- a numeric vector with mean reaction times in a lexical decision task
Source (http://elexicon.wustl.edu/WordStart.asp)

Data from Natalya Levshina's `RLing` package available (here)[https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/ELP.csv]

### 2.0 Read the data from file to the variable `elp`.
```{r 2.0}
elp <- read.csv('https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/ELP.csv')
head(elp)
```

### 2.1 Which two variables have the highest Pearson's correlaton value?
```{r 2.1}
cor(elp$Length, elp$SUBTLWF, method='pearson')
cor(elp$Length, elp$Mean_RT, method='pearson')
cor(elp$SUBTLWF, elp$Mean_RT, method='pearson')
### Length and Mean_RT appeared to have the highest Pearson's correlation value (~0.5276826).
```

### 2.2 Group your data by parts of speech and make a scatterplot of SUBTLWF and Mean_RT.
```{r 2.2}
elp %>%
ggplot(aes(x = SUBTLWF, y = Mean_RT, colour = Length)) + xlim(0, 200) + geom_point() + scale_color_continuous(low = "lightblue", high = "red") + facet_wrap( ~ POS) + labs(caption = 'data from (Balota et al. 2007)') + scale_x_log10()
```

### 2.3 Use the linear regression model to predict `Mean_RT` by `log(SUBTLWF)` and `POS`.
#### 2.3.1 Provide the result regression formula
```{r 2.3.1}
linearMod <- lm(Mean_RT ~ log(SUBTLWF) + POS, data=elp)
summary(linearMod)
```

#### 2.3.2 Provide the adjusted R$^2$
```{r 2.3.2}
### The adjusted R-squared is equal to 0.3385.
```

#### 2.3.3 Add the regression line in the scatterplot.
```{r 2.3.3}
elp %>%
ggplot(aes(x = log(SUBTLWF), y = Mean_RT, colour = Length)) + geom_point() + scale_color_continuous(low = "skyblue", high = "red") + labs(caption = 'data from (Balota et al. 2007)') + xlab("log(SUBTLWF)") + geom_smooth(method = "lm", se = FALSE, colour = 'black')
```

### 2.4 Use the mixed-efects model to predict `Mean_RT` by `log(SUBTLWF)` using POS intercept as a random effect

#### 2.4.1 Provide the fixed effects formula
```{r 2.4.1}
fixed_effects <- lmer(Mean_RT~log(SUBTLWF) + (1|POS), data=elp)
summary(fixed_effects)
```

#### 2.4.2 Provide the variance for intercept argument for `POS` random effects
```{r 2.4.2}
###The variance is equal to 414.4
```

#### 2.4.3 Add the regression line to the scatterplot
```{r 2.4.3}
elp %>%
ggplot(aes(x = log(SUBTLWF), y = Mean_RT, colour = POS)) + geom_point() + labs(caption = 'data from (Balota et al. 2007)') + xlab("log(SUBTLWF)") + geom_smooth(method = "lm", se = FALSE, colour = 'black') + facet_wrap( ~ POS) + theme(legend.position="none")
```

## 3. Dutch causative constructions

This is a data set with examples of two Dutch periphrastic causatives extracted from newspaper corpora.

The data frame includes 100 observations on the following 7 variables:

* Cx -- a factor with levels doen_V and laten_V
* CrSem -- a factor that contains the semantic class of the Causer with levels Anim (animate) and Inanim (inanimate).
* CeSem -- a factor that describes the semantic class of the Causee with levels Anim (animate) and Inanim (inanimate).
* CdEv -- a factor that describes the semantic domain of the caused event expressed by the Effected Predicate. The levels are Ment (mental), Phys (physical) and Soc (social).
* Neg -- a factor with levels No (absence of negation) and Yes (presence of negation).
* Coref -- a factor with levels No (no coreferentiality) and Yes (coreferentiality).
* Poss -- a factor with levels No (no overt expression of possession) Yes (overt expression of possession)

Data from Natalya Levshina's `RLing` package available (here)[https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv]

### 3.0 Read the data from file to the variable `d_caus`.
```{r 3.0}
d_caus <- read.csv('https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv')
head(d_caus)
```

### 3.1 We are going to test whether the association between `Aux` and other categorical variables (`Aux` ~ `CrSem`, `Aux` ~ `CeSem`, etc) is statistically significant. The assiciation with which variable should be analysed using Fisher's Exact Test and not using Pearson's Chi-squared Test? Is this association statistically significant?
```{r 3.1}
chisq.test(d_caus$Aux, d_caus$CrSem)
chisq.test(d_caus$Aux, d_caus$CeSem)
chisq.test(d_caus$Aux, d_caus$CdEvSem)
chisq.test(d_caus$Aux, d_caus$CeSynt)
fisher.test(d_caus$Aux, d_caus$CeSynt)
chisq.test(d_caus$Aux, d_caus$EPTrans)
chisq.test(d_caus$Aux, d_caus$Country)
chisq.test(d_caus$Aux, d_caus$Domain)
### The results show us that Pearson's chi-square test is an inappropriate way to test the association between Auc and CeSynt. The reason for it is that the expected frequences seem to be rather low. Nevertheless, taking into account the p-value of Fisher's test, this association can still be statistically significant.
```

### 3.2. Test the hypothesis that `Aux` and `EPTrans` are not independent with the help of Pearson's Chi-squared Test. 
```{r 3.2}
aux_table <- table(d_caus$Aux,d_caus$EPTrans)
chisq.test(aux_table)
### Here we have a fairly low p-value, that indicates an accordingly high chance of having an association between the two columns.
```

### 3.3 Provide expected values for Pearson's Chi-squared Test of `Aux` and `EPTrans` variables.
```{r 3.3}
chisq.test(d_caus$Aux, d_caus$Country)$expected
```

### 3.4. Calculate the odds ratio.
```{r 3.4}
oddsratio(d_caus$Aux~d_caus$EPTrans)
### The ratio is equal to 0.9578785.
```

### 3.5 Calculate effect size for this test using Cramer's V (phi).
```{r 3.5}
### Using the formula from Lab5 (V = sqrt(X-squared / [n_obs * (min(ncols, nrows)-1“ 1)])):
V <- sqrt(14.307/500*(min(500,2)-1))
V
### The Carmer's V is equal to 0.1691567.
```

### 3.6. Report the results of independence test using the following template:
```{r 3.6}
### A significant association between variables Aux and Country was found (p = 0.0001127 < 0.001). The odds of AUX are 2,6 times higher in Belgian Dutch than in Netherlands Dutch. The effect size is rather small (Cramer's V = 0.178).
```

### 3.7 Visualize the distribution using mosaic plot.
Use `mosaic()` function from `vcd` library.
```{r 3.7}
vcd::mosaic(~ Aux + EPTrans, data=d_caus, shade=TRUE, legend=TRUE)
```

Below is an example of how to use mosaic() with three variables.
```{r 3.7.1}
# mosaic(~ Aux + CrSem + Country, data=d_caus, shade=TRUE, legend=TRUE)
```

### 3.8 Why is it not recommended to run multiple Chisq tests of independence on different variables within your dataset whithout adjusting for the multiplicity? (i.e. just testing all the pairs of variables one by one)  
```{r 3.8}
### The answer is simple: it is done in such a way in order not to increase the chance to reject the correct H0-hypothesis.
```

### 3.9 Provide a short text (300 words) describing the hypothesis on this study and the results of your analysis.
```{r 3.9}
### In the course of the study different variables were analyzed (namely: country, transitivity, animacy and semantic variables), as well as the way the variables influence the people's choiñe between doen or laten.
### H0: There is no strong correlation between auxiliary verbs and their transitivity.
### H1: On the contrary, the correlation between the two variables exists.
### Analyzing the chi-square test results, we recieved p=0.0001553<0.05, what gives us reason to reject H0 in favour for H1. However, the odds ratio is fairly small and suggests that two variables could not be strongly linked.
```
